---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# GeneExprLupusMx

## Explorando el lupus en México: datos abiertos y análisis reproducibles en R

🎯 El objetivo de `GeneExprLupusMx` es proporcionar un paquete de datos en R, bilingüe y reproducible, que integre perfiles transcriptómicos y metadatos clínicos de mujeres mexicanas con **lupus eritematoso sistémico (SLE) y controles sanos**. Al combinar datos de expresión génica de *monocitos, células dendríticas derivadas de monocitos (moDCs) y células dendríticas tolerogénicas (tolDCs)*, el paquete permite:

- **Uso educativo:** servir como recurso didáctico en bioinformática, inmunología y ciencias biomédicas.

- **Análisis comparativo:** apoyar la exploración de redes de regulación genética y estudios de expresión diferencial.

- **Flujos reproducibles:** ofrecer funciones auxiliares y scripts para replicar figuras y análisis del estudio original.

- **Integración global:** hacer visibles y utilizables los conjuntos de datos generados localmente en contextos internacionales de investigación.

- **Divulgación científica:** promover la ciencia abierta y la difusión de datos biomédicos de América Latina.

------------

## Exploring lupus in Mexico: open data and reproducible analysis in R”

The goal of `GeneExprLupusMx` is to provide a bilingual, reproducible R data package that integrates transcriptomic profiles and clinical metadata from Mexican women with **systemic lupus erythematosus (SLE) and healthy controls**. By combining gene expression data from *monocytes, monocyte-derived dendritic cells (moDCs), and tolerogenic dendritic cells (tolDCs)*, the package enables:

- **Educational use:** serving as a teaching resource in bioinformatics, immunology, and biomedical sciences.

- **Comparative analysis:** supporting exploration of gene regulation networks and differential expression studies.

- **Reproducible workflows:** offering auxiliary functions and scripts to replicate figures and analyses from the original study.

- **Global integration:** making locally generated datasets visible and usable in international research contexts.

- **Scientific outreach:** promoting open science and the dissemination of biomedical data from Latin America.

## Instalación / Installation

Puedes instalar la versión en desarrollo del paquete `GeneExprLupusMx` desde GitHub con: **/** You can install the development version of `GeneExprLupusMx` from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("EveliaCoss/GeneExprLupusMx")
```

## Sobre los datos / About the data

Derivado del estudio **"Gene expression profiling of dendritic cell tolerance dysfunction in women with Systemic lupus erythematosus"**, este recurso integra datos de expresión génica y metadatos clínicos y experimentales de mujeres mexicanas con lupus y controles sanas, en los tres tipos celulares mencionados: *monocitos, células dendríticas derivadas de monocitos (moDCs) y células dendríticas tolerogénicas (tolDCs).* **/** Derived from the study **"Gene expression profiling of dendritic cell tolerance dysfunction in women with Systemic lupus erythematosus"**, this resource integrates gene expression data and clinical and experimental metadata from Mexican women with lupus and healthy controls, across the three mentioned cell types: *monocytes, monocyte-derived dendritic cells (moDCs), and tolerogenic dendritic cells (tolDCs).*

El paquete `GeneExprLupusMx` contiene 4 datasets. **/** The `GeneExprLupusMx` package contains 4 datasets:

- `metadata_cleaned`: 
- `metadata`: Archivo suplementario
- `filtered_txi`: Datos sobre expresión diferencial
- `Redes de regulación`:

Cargar datasets **/** Load datasets:

```{r example}
library(GeneExprLupusMx)
data(package = 'GeneExprLupusMx')
```

One is called `metadata_cleaned`, and is a simplified version of the raw data; see `?metadata_cleaned` for more info:

```{r metadata}
summary(metadata_cleaned)
```

The second dataset is penguins_raw, and contains all the variables and original names as downloaded; see ?penguins_raw for more info.

```{r filtered_txi}
str(filtered_txi)
```

Both datasets contain data for 344 penguins. There are 3 different species of penguins in this dataset, collected from 3 islands in the Palmer Archipelago, Antarctica.

## Ejemplos / Examples

```{r}
library(tidyverse)
metadata_cleaned %>% 
  count(Group, Cell_type)
```

Penguins are fun to visualize! For example:



## Cómo citar este paquete / How to cite this package

To cite the palmerpenguins package, please use:

```{r}
citation("GeneExprLupusMx")
```

## Additional data use information

donde vive los datos crudos

## References

Data originally published in:

- Hernández-Ledesma AL, Coss-Navarrete EL, Salazar-Magaña S, Ramírez-Espinosa D, Tinajero-Nieto L, Torres-Valdez E, Peña-Ayala AH, Félix-Rodriguez G, Frontana-Vázquez G, García Sotelo JS, Thomas-Chollier M, Trynka G, Rosetti F, Fernandez-Valverde SL, Gutiérrez-Arcelus M, Alpízar-Rodríguez D and Medina-Rivera A (2026) Gene expression profiling of dendritic cell tolerance dysfunction in women with systemic lupus erythematosus. Front. Immunol. 17:1771959. doi: 10.3389/fimmu.2026.1771959
